Demystifying the Computer Science PhD Admission\\in US Universities\\{\large A Guide for International Students

### 2023-11-08 demystify.tex
updates

So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines and should be 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

\chapter{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a PhD in CS in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

\mainmatter
\chapter{Should You Apply to a PhD program in CS in the US?}\label{sec:should}

>>>>

So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

\chapter{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a CS PhD in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item GRE \emph{is not} required by most schools (\S\ref{sec:grades}).
    \item Grades are important, but only if your school is well-known (\S\ref{sec:grades}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

\mainmatter
\chapter{Should You Apply to a CS PhD Program in the US?}\label{sec:should}
